---
title: "HIF-1 Dataset"
author: "Rongo Lab"
date: "2/27/2022"
output:
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Reading Count Data into R and Omit NA

Determine count data file path.Read data from .csv file and store into *rawdataCR*. Data is of type data frame.

Remove all NA values from data frame. Label data frame columns.

```{r Reading Count Data, echo=TRUE}
path <- file.path("~","desktop","GeneCounts_HIF1_Dataset.csv")
rawdataCR <- read.csv(path,stringsAsFactors = FALSE,header = TRUE)
ctdataCR <- na.omit(rawdataCR)
```

Load EdgeR.

Label data frame columns (N2cgc_1, N2cgc_2, N2cgc_3, N2cgc_4, egl9_B, egl9_C, egl9_H, egl9_I, hif1_B, hif1_C, hif1_H, hif1_I, egl9hif1_B, egl9hif1_C, egl9hif1_H, egl9hif1_I).

Check data frame column labels by examining top six rows.

```{r}
group <- c("WBID","N2cgc_1","N2cgc_2","N2cgc_3","N2cgc_4","egl9_B","egl9_C", "egl9_H", "egl9_I", "hif1_B",	"hif1_C",	"hif1_H",	"hif1_I",	"egl9hif1_B",	"egl9hif1_C",	"egl9hif1_H",	"egl9hif1_I")
colnames(ctdataCR) <- group
head(ctdataCR)

```

The EdgeR package stores the data in a list object called *DGEList*.

Name data set groups and create list object. This is *lstC*.

This list can be treated like a list or subsetted like a matrix.

Subset DGEList object passing only read counts from the data frame *rawdataCR*. This is *lstC*.

The Worm Base gene identifier is the first column of the data frame *rawdataCR*. The **genes** element in DGEList is the Worm Base gene identifier used in the data frame *rawdataCR*.

Examine first six rows of data set in list object and check dimensions of *lstC*. Output is: 47096 rows and 16 columns.

```{r}
group <- c("N2cgc_1","N2cgc_2","N2cgc_3","N2cgc_4","egl9_B","egl9_C", "egl9_H", "egl9_I", "hif1_B",	"hif1_C",	"hif1_H",	"hif1_I",	"egl9hif1_B",	"egl9hif1_C",	"egl9hif1_H",	"egl9hif1_I")
lstC<-DGEList(genes=ctdataCR[,1],counts = ctdataCR[,2:17],group = group)
head(lstC)
dim(lstC)
```

## Filtering: Removing low read counts

Genes that have a low count have to be removed. This makes sense biologically because a gene has to have a minimal count to be translated or to have any biological significance. Statistically, a gene has to have a minimal count to provide sufficient evidence that a gene is differentially expressed.

Filtering is based on counts per million (CPM) to avoid favoring of genes that are expressed in large libraries over those expressed in smaller libraries. The cpm values are *lstCPM*.

```{r}
lstCPM <- cpm(lstC)
```

Here we keep counts that are greater than or equal to 0.5 CPM. The counts that are greater than 0.5 are *thresh*.

This cut-off of 0.5 is equal to 10/*L* where *L* is the minimum library size that is considered acceptable (*L* =20 million in this data analysis).

Examining the first six rows of *thresh* shows a logical matrix of TRUEs and FALSEs.

```{r}
thresh <- lstCPM > 0.5
head(thresh)
```

Find total number of TRUE values in each row of the matrix

```{r}
table(rowSums(thresh))
```

Keep genes with CPM values that are greater than 0.5 in at least 4 libraries since there are 4 replicates per group. Keep genes that have at least 4 TRUEs in each row of thresh.

```{r}
lstCPM.keep <- rowSums(thresh)>=4
#summary(keep)
```

Subset rows of count data in *lstC* to keep genes that are highly expressed.

```{r}
lstC.keep <- lstC[lstCPM.keep, ]
dim(lstC)
```

```{r}
dim(lstC.keep)
```

Recompute the library sizes of the highly expressed genes data set

```{r}
lstC.keep$samples$lib.size <- colSums(lstC.keep$counts)
```

Apply TMM normalization to the data set to account for compositional differences between the libraries.

```{r}
lstC.keep <- calcNormFactors(lstC.keep)
head(lstC.keep$samples)
tail(lstC.keep$samples)
```

##Examine Data

Plot a multidimensional scaling analysis to examine data for outliers and/or any relationships in the data.

```{r}
mds <- plotMDS.DGEList(lstC.keep, xlab ="Dimension 1",ylab ="Dimension 2",asp = 1,cex = 0.8,col=c(rep("blue",4),rep("darkgreen",4),rep("red",4),rep("black",4)))
plot(mds,xlab ="Dimension 1",ylab ="Dimension 2",asp = 6/9,cex = 0.8,col=c(rep("blue",4),rep("darkgreen",4),rep("red",4),rep("black",4)))


```

```{r}
plotMDS.DGEList(lstC.keep, xlim = c(-3,3), ylim = c(-2,2))
```

```{r pressure, echo=FALSE}
#plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
